---
title: "MiniProject"
author: "PetraGuy"
date: "23 November 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
clear the workspace
```{r}

rm(list = ls())
cat("\014")
```
Get the data , enter input CSV file name here, for data in data dicerctory
```{r}
inputfile = 'SorariaCompact1.csv'
Dataname = strsplit(inputfile, "\\.")[[1]][[1]]
fullfile = paste("../Data",inputfile,sep = '/')
Data = read.csv(fullfile)
```

Do some median imputations if necessary
```{r}
median_replace1 = function(x){
  ifelse(is.na(x), median(x,na.rm = TRUE), x)
 }

median_replace2 = function(x){
  apply(x,2,median_replace1)
}

Imputed_list = lapply(split.data.frame(Data[,2:12], Data$Species), FUN = median_replace2)
```

The imputed dataframe is a list with species as the elements, the following sticks it back together with a different name so both optiona are available

```{r}
temp = do.call(rbind, Imputed_list)
Imputed_df = cbind(Data[1], temp)
```

Its worth creating a scaled dataframe too for some of the analysis and ploting
```{r}
Scaled_df = scale(Imputed_df[-1])
Scaled_df = cbind(Data[1], Scaled_df)
```


PLOTTING

Execute the first section to save the data, or go straight to plotting to view graphs in document
```{r}
out = paste(Dataname, "pairs", sep = "_")
out= paste('../Results',out,sep = '/')
out = paste(out,'.pdf')

```
Pairs Plots, unhash pdf to save data to file. Not all pairs plotted beacuse there are so many the plot becomes too difficult to read and therefore uninformative. The pairs plot

```{r}
#pdf(out)
reduced_df = Imputed_df[-c(4,5,6,7,8,9,12)]
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}
pairs(reduced_df[-1], lower.panel = panel.smooth, upper.panel = panel.cor, col = Imputed_df$Species)
mtext(Dataname, side = 1)
```

Repeat the pairs plots using ggplot

```{r}
library(GGally)
ggpairs(data = Imputed_df[,2:6])

ggpairs(Imputed_df[,7:12])
```



Look at scatter plots by species of main characters, run the following to get new filename if saving data
```{r}
out = paste(Dataname, "scatter", sep = "_")
out= paste('../Results',out,sep = '/')
out = paste(out,'.pdf')

```

Scatter plots, unhash pdf to save data to file

```{r}
library(ggplot2)
library(gridExtra)
#pdf(out)
plot1= ggplot(Imputed_df, aes(x= LeafLength, y = LeafWidth, col = Species))+
  geom_point()
plot2 = ggplot(Imputed_df, aes(x= FruitLength, y = FruitWidth, col = Species))+
  geom_point()

grid.arrange(plot1,plot2)
```

Box plots
```{r}
out = paste(Dataname, "box", sep = "_")
out= paste('../Results',out,sep = '/')
out = paste(out,'.pdf')
```


```{r}
#pdf(out)
col_names = colnames(Data)
col_names = col_names[-1]
plotlist = list()
library(grid)
for (i in seq_along(col_names)){
  plot = ggplot(Imputed_df, (aes_string(x='Species', y=col_names[i])))+
                 geom_boxplot()
  plotlist = rbind(plotlist,  plot)
  grid.draw(plot)
}

```

Try ggplot option for boxplots, but they'll need scaling because facet wrap will use same scale for all

```{r}
library(reshape2)
melted = melt(Scaled_df)

ggplot(data = melted) +  geom_boxplot(aes(x=Species,y=value, fill = Species)) +   facet_wrap(~variable) +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```

Th scatter plots and pairs plots indicate that only leaf length, leaf width and widest point are correlated with a correlation coefficient of 0.9. Two of these variables might then be removed from the analysis.

Lets look at anovas for each feature

```{r}

col_names = colnames(Data)
col_names = col_names[-1]
anova_data = data_frame()
model = list()
for (i in seq_along(col_names)){
    m = lm(formula = paste(col_names[i], "~ Species"), data =  Imputed_df)
    ANOVA <- anova(m)
    r2 <- summary(m)$r.squared
    data = data.frame(col_names[i], ANOVA[[4]][1], ANOVA[[5]][1], r2)
    anova_data = rbind(anova_data,data)
    
    
  }
names(anova_data) = c("Feature","F", "p", "r2")
print(anova_data)
```

