---
title: "DBH simulations"
author: "PetraGuy"
date: "28 January 2018"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
#clear the workspace
rm(list = ls())
cat("\014")
```

The purpose of this document is to create some simulated dbh frequency distributions, that mimic the range of distributions that might be found in the sites under consideration. A statistic must be derived from these distributions which reflects the succession of a plot since an opening event. This statistic must be easily derived from actual dbh data.

Disitributions used to model dbh include the two and three paramater Weibull disribution and the logit-logistic.
Two paramater Weibull:

$$ f(x) = {\kappa\lambda}^{\kappa}{x}^{-(\kappa-1)}e^{-{({\lambda}{x})}^{\kappa}}$$
Three paramater Weibull:

$$ f(x) = \frac{\beta}\eta\frac{x-\gamma}{\eta} e^{-(\frac{x-\gamma}{\eta})^{\beta}} $$
And the logit-logistic


The summary statistics from the 103 woodlands indicates that the simulation should have a range for x of 1 to 32. The maximum frequency was 372 with the first dbh class being the mode. 
The simulation aims tp approximate the frequncies seen in the woodland survey, whilst being a general exploration
```{r}
#lets try the 2 parameter weibull. It doesn't have to be a perfect fit to data - just has to simulate some 
#realistic looking distributions
# first need to look at the effects of the 2 parameters
#fix beta and n=eta, vary gamma 
i = seq(from=0.1, to =2, by = 0.1)

weibull = function(x,l=1,k=5){
term1 = k*l^x
term2 = x/(x^k)
term3 = exp(-1*(l*x)^k)
f = term1*term2*term3
return(f)
}

f = vector()
for (i in 1:200){
  x = 0.01*i
  f[i]= weibull(x)
}
x = c(1:200)
x = x *0.01

plot(x,f, type = "l")

```



