---
title: "DBH simulations"
author: "PetraGuy"
date: "28 January 2018"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
#clear the workspace
rm(list = ls())
cat("\014")
```

The purpose of this document is to create some simulated dbh frequency distributions, that mimic the range of distributions that might be found in the sites under consideration. A statistic must be derived from these distributions which reflects the succession of a plot since an opening event. This statistic must be easily derived from actual dbh data.

A distribution that could be used to create the simulated data is the three paramater Weibull disribution given by

$$ f(x) = \frac{\beta}\eta\frac{x-\gamma}{\eta} e^{-(\frac{x-\gamma}{\eta})^{\beta}} $$

The summary statistics from the 103 woodlands indicates that the simulation should have a range for x of 1 to 32. The maximum frequency was 372 with the first dbh class being the mode. 
The simulation aims tp approximate the frequncies seen in the woodland survey, whilst being a general exploration
```{r}
# first need to look at the effects of the 3 parameters
#fix beta and n=eta, vary gamma 
x = c(1:32)


```



