---
title: "ZetaDiversityExample"
author: "PetraGuy"
date: "28 November 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
graphics.off()
```

I think this might be the way Zeta Diversity works??

Imagine we have 4 sites with a set of species in each site. The species are represented by numbers 1-4.
These simplistic sites meean we can easily see which species are shared and what to expect. (PS I created a list too because I'm not sure what objects will be needed for the calculations)

```{r}
Site1 =  c(1:4)
Site2 = c(2:5)
Site3 = c(3:6)
Site4 = c(4:7)
Sites = cbind(Site1,Site2,Site3,Site4)
SiteList = list(Site1,Site2,Site3,Site4)

print(Sites)

```

How many species occur in all 4 sites - whats the intersection of all 4 sites?

```{r}
All4 = Reduce(intersect, list(Site1,Site2,Site3,Site4))
richness = length(unique(All4))
print(c("richness", richness))
z4 = (All4)
print(c("z4", z4))
```

How many species occur in 3 out of  4 sites. Need to find all the combinations of 3 from 4, (combs) then use the intersection as above. The columns of combs are the different combinaions of 4 choose 3. So selecting each column (combs[,i]) in turn gives a unique way that the sites must be combined. Then each element of that combination (combo[j]) gives an index for the columns of the Sites - i.e, a single site

```{r}
combs = combn(4,3)
richness = vector()
for (i in 1:ncol(combs)){
  combo = combs[,i]
   x =  Reduce(intersect, list(Sites[,combo[3]],Sites[,combo[2]],Sites[,combo[1]]))
  print(x)
  richness[i] = length(unique(x))
  
} 
print(c("richness", richness))
z3 = mean(richness)
print(c("z3", zeta))
 
```
Which species occur in 2 of the 4 sites

```{r}
combs = combn(4,2)
richness = vector()
for (i in 1:ncol(combs)){
  combo = combs[,i]
   x =  Reduce(intersect, list(Sites[,combo[1]],Sites[,combo[2]]))
  print(x)
  richness[i] = length(unique(x))
  
} 
print(c("richness", richness))
z2 = mean(richness)
print(c("z2",z2))

```
What's the avergae number of species across all sites

```{r}
library(reshape)
species = melt(Sites)
richness = length(unique(species$value))
z1 = mean(richness)/4
print(c("z1", z1))
```
The plot would be zeta vs number of sites

```{r}
x = (1:4)
y = c(z1,z2,z3,z4)
plot(x,y)
```
This is simplistic, ugly and clunky, but I was mainly trying to see if I had the right idea for what zeta diversity is. If this is correct, Hui et al have written an R package anyway, so I wont this!

The next step, I think, is to extrapolate the above graph to get the estimate for species area??


  print(x)
  richness[i] = length(unique(x))
  
} 
print(c("richness", richness))
z2 = mean(richness)
print(c("z2",z2))

```
What's the avergae number of species across all sites

```{r}
library(reshape)
species = melt(Sites)
richness = length(unique(species$value))
z1 = mean(richness)/4
print(c("z1", z1))
```
